---
  title: "Programming with data: the R language"
  author: Robert Castelo
  output:
    xaringan::moon_reader:
      nature:
        beforeInit: js/macros.js
      self_contained: true
      css: [default, default-fonts, ninjutsu, css/upf.css]
      seal: false
---

```{r setup, include=FALSE, cache=FALSE}
options(htmltools.dir.version = FALSE, width=80)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache=TRUE)

## from http://yihui.name/knitr/hooks#chunk_hooks
knitr::knit_hooks$set(small.mar=function(before, options, envir) {
                                              if (before) par(mar=c(4, 5, 1, 1)) ## smaller margin on top and right
})
```

class: title-slide, middle, center

# Data manipulation and plotting with R

## Mireia Ramos
[mireia.ramos@upf.edu](mailto:mireia.ramos@upf.edu)
### Dept. of Experimental and Health Sciences
### Universitat Pompeu Fabra

<br>

## Fundamentals of Computational Biology
### BSc on Human Biology
### UPF School of Health and Life Sciences
### Academic Year 2021-2022

---
class: center, middle, inverse

# Control flow statements

---

## Control flow statements

* Sequentially evaluated statements may be written one below another using
  newlines or in the same line separated by semicolons (`;`):
  <pre>
  i <- 1 ; j <- 0
  </pre>
* Conditional and looping statements have the condition surrounded by
  parentheses `( )` and the associated code by braces `{ }`, except if it
  consists of a single line.
  <pre>
  if (i > 0)
      j <- i * 10

  while (i < j) {
      i <- i + 1
      print(i)
  }
  for (i in 1:10) {
      j <- j + i
      print(j)
  }
  </pre>

---

## Interpretation and vectorization

* The R language is interpreted, vectorized and includes implicit memory
  management.  
  &nbsp;&nbsp;
* Looping statements are slow in R. The following example would be a (wrong)
  approach in R following a classical programming style:
  <pre>
  x <- 1:100
  logsum <- 0
  for (i in 1:length(x)) {
      logsum <- logsum + log(x[i])
  }
  </pre>
* This would be the vectorized (right) approach in R:
  <pre>
  logsum <- sum(log(x))
  </pre>

---

## `apply()` functions

* To traverse the data, R uses the apply functions.
* `apply()`.
* `lapply()`.
* `sapply()`.
* `tapply()`

---

## Concluding remarks

* Control flow statements can conditionally execute code surrounded by braces.
  Identation is only necessary for readability.  
  &nbsp;&nbsp;
* Vectorized operations are much faster than loops.  
  &nbsp;&nbsp;
  
---
class: center, middle, inverse

# Data manipulation

---

## Add columns to data frames

---

## The recycling rule

* When performing operations between vectors of different length, R
  follows a specific policy called the _recycling_ rule, by which the
  shorter vector is _recycled_.  
  &nbsp;&nbsp;
* This rule is straightforward when the length of the longer vector is multiple
  of the length of the shorter vector:
  <pre>
  > 2 * 1:4
  [1] 2 4 6 8
  > 2:3 * 1:4
  [1] 2  6  6  12
  </pre>
* When the length of the longer vector is not multiple of the length of the
  shorter vector, R issues a warning:
  <pre>
  > 2:3 * 1:5
  [1] 2  6  6 12 10
  Warning message:
  In 2:3 * 1:5 :
    longer object length is not a multiple of shorter object length
  </pre>

--- 

## Sorting and ordering data

* `sort()`
* `order()`

---

## Combining and Merging datasets

* Datasets with the same shape can be combined using:
  * `cbind()` to add columns. 
  * `rbind()` to add rows.
* Datasets with the same shape can be combined using: 
  * `merge()`, using common columns and rows (join)
  * `match()`, returns the matching positions of two vectors. 

--- 

## The "join" concept

* _Natural join_. Keep only rows that match in both data frames. 
* _Full outer join_. Keep all rows from both data frames.
* _Left outer join_. Include all rows from data frame `x` and only those from `y` that match.
* _Right outer join_. Include all rows from data frame `y` and only those from `x` that match.

---

## Splitting data

* `split()`

---
class: center, middle, inverse

# Plotting with R

---
## Generic plot types in R

* The basic plotting function in R is `plot()`.

```{r, out.width="50%"}
plot(x=iris$Sepal.Length, y=iris$Sepal.Width, type="p")
```






